**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**

### 2021-05-27 Manuscripts/Manuscript by chapters -v2/06-Appendix.Rmd
Making proposed changes in the manuscript

```{r, responses-Appendix,include=TRUE,message=FALSE,echo=TRUE}
set.seed(1)
####################Section for calculations######################################
################################################################################

@@

#plotting simulated datapoints and smooth term
p14<-ggplot(data=dat,
            aes(x=time,y=y))+
  geom_point(size=sz+1)+
  scale_color_aaas()+
  labs(y='Simulated \n response')+
  geom_line(data=smooth,
            aes(x=time,
                y=smooth),
            color="#6C581DFF",
            size=sz+1)+
  theme_classic()

@@

From the `summary` function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained. The link function is 'identity' as the model does not make any transformation on the predictors. The 'significance of smooth terms' _p-value_ indicates if each smooth is adding significance to the model. Here, the _p-value_ is low but we have seen that there are issues with the model from the previous outputs. Finally, the 'deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to ~12%.

@@

Diagnostics for this model indicate that the k-index is still below 1 (0.32 from `gam.check`), and that the residuals are still not following a normal distribution (Figure  \@ref(fig:second-GAM-diag)). Moreover, the smooths (plotted via the `draw()` function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data. From `summary()`, the deviance explained by the model is ~43%.

@@

The resulting model is`m1`, which is the model fitted in the main manuscript. By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group . From `gam.check`, the k-index is now at an acceptable value (~1.02), and `summary` now indicates that the model is able to capture 87% of the variance data.

>>>>

```{r, responses-Appendix,include=TRUE,message=FALSE,echo=TRUE}
set.seed(1)
##########Section for calculations###########

@@

#plotting simulated datapoints and smooth term
p14<-ggplot(data=dat,
            aes(x=time,y=y))+
  geom_point(size=sz+1)+
  labs(y='Simulated \n response')+
  geom_line(data=smooth,
            aes(x=time,
                y=smooth),
            color="#6C581DFF",
            size=sz+1)+
  theme_classic()

@@

From the `summary` function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained. The link function is 'identity' as the model does not make any transformation on the predictors. The 'significance of smooth terms' _p-value_ indicates if each smooth is adding significance to the model. Here, the _p-value_ is low but we have seen that there are issues with the model from the previous outputs. Finally, the 'deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to $\approx$12%.

@@

Diagnostics for this model indicate that the k-index is still below 1 (0.32 from `gam.check`), and that the residuals are still not following a normal distribution (Figure  \@ref(fig:second-GAM-diag)). Moreover, the smooths (plotted via the `draw()` function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data. From `summary()`, the deviance explained by the model is $\approx$43%.

@@

The resulting model is`m1`, which is the model fitted in the main manuscript. By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group . From `gam.check`, the k-index is now at an acceptable value ($\approx$1.02), and `summary` now indicates that the model is able to capture 87% of the variance data.

### 2021-05-27 Manuscripts/Manuscript by chapters -v2/Full_document.Rmd
Making proposed changes in the manuscript

In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, a _linear mixed model_ (LMEM). Although LMEMs are less restrictive that rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear. In contrast, generalized additive models (GAMs), are a class of models that relax the linearity assumption and allow the data to determine the fit of the model while permitting missing observations and different correlation structures, thereby being an excellent choice to analyze non-linear longitudinal data. This paper summarizes the limitations of LMEMs and rm-ANOVA, presents the basic theory of GAMs, and demonstrates their implementation in $\textsf{R}$ via the package _mgcv_ using simulated data that follows longitudinal trends reported in biomedical literature. To promote reproducibility in biomedical research, the code and data used to generate this paper are available at:________.

>>>>

In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, a _linear mixed model_ (LMEM). Although LMEMs are less restrictive that rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear. In contrast, generalized additive models (GAMs), are a class of models that relax the linearity assumption and allow the data to determine the fit of the model while permitting missing observations and different correlation structures, thereby being an excellent choice to analyze non-linear longitudinal data. This paper summarizes the limitations of rm-ANOVA and LEMEs, presents the basic theory of GAMs, and demonstrates their implementation in $\textsf{R}$ via the package _mgcv_ using simulated data that follows longitudinal trends reported in biomedical literature. To promote reproducibility in biomedical research, the code and data used to generate this paper are available at:________.
